\documentclass{article}

\setlength{\headsep}{0.75 in}
\setlength{\parindent}{0 in}
\setlength{\parskip}{0.1 in}

%=====================================================
% Add PACKAGES Here (You typically would not need to):
%=====================================================

\usepackage{xcolor}
\usepackage[margin=1in]{geometry} % Used for page margins, often replaces fullpage
\usepackage{amsmath,amsthm,amssymb} % amssymb added for common math symbols

\usepackage{enumitem} % For custom list environments

\usepackage{tcolorbox}
% \usepackage{fullpage} % Geometry handles this better
\usepackage[parfill]{parskip} % For paragraph spacing
\usepackage{hyperref}
\usepackage[nameinlink, capitalise]{cleveref} % For smart referencing

\usepackage{algorithm2e} % For algorithms
\usepackage{tikz} % For drawing diagrams
\usepackage{siunitx} % For units
\usepackage{graphicx}

\usetikzlibrary{arrows.meta, positioning}

\crefname{algocf}{Algorithm}{Algorithms}

\RestyleAlgo{ruled} % Style for algorithms

\definecolor{linkc}{rgb}{0.1, 0.5, 0.7}
\definecolor{citec}{rgb}{0.6, 0.3, 0.7}
\definecolor{urlc}{rgb}{0.5, 0.1, 0.2}
\hypersetup{
    colorlinks=true,
    linkcolor=linkc,
    citecolor=citec,
    urlcolor=urlc
}

\renewcommand{\d}{\,\mathrm{d}} % Differential d


%=====================================================
% Ignore This Part (But Do NOT Delete It:)
%=====================================================

\theoremstyle{definition}
\newtheorem{problem}{Problem}
\newtheorem*{fun}{Fun with Algorithms}
\newtheorem*{challenge}{Challenge Yourself}
\def\fline{\rule{0.75\linewidth}{0.5pt}}
\newcommand{\finishline}{\begin{center}\fline\end{center}}
\newtheorem*{solution*}{Solution}
\newenvironment{solution}{\begin{solution*}}{{\finishline} \end{solution*}}
\newcommand{\grade}[1]{\hfill{\textbf{($\mathbf{#1}$ points)}}}
\newcommand{\thisdate}{\today}
\newcommand{\thissemester}{\textbf{Rutgers: Spring 2026}}
\newcommand{\thiscourse}{CS 344: Algorithms} 
\newcommand{\thishomework}{Number} 
\newcommand{\thisname}{Name} 
\newcommand{\thisextension}{Yes/No} 

\headheight 40pt              
\headsep 10pt


\newcommand{\thisheading}{
   \noindent
   \begin{center}
   \framebox{
      \vbox{\vspace{2mm}
    \hbox to 6.28in { \textbf{\thiscourse \hfill \thissemester} }
       \vspace{4mm}
       \hbox to 6.28in { {\Large \hfill Homework \#\thishomework \hfill} }
       \vspace{2mm}
         \hbox to 6.28in { { \hfill \hfill} }
       \vspace{2mm}
       \hbox to 6.28in { \emph{Name(s): \thisname \hfill }}
      \vspace{2mm}}
      }
   \end{center}
   \bigskip
}

%=====================================================
% Some useful MACROS (you can define your own in the same exact way also)
%=====================================================


\newcommand{\ceil}[1]{{\left\lceil{#1}\right\rceil}}
\newcommand{\floor}[1]{{\left\lfloor{#1}\right\rfloor}}
\newcommand{\prob}[1]{\Pr\paren{#1}}
\newcommand{\expect}[1]{\Exp\bracket{#1}}
\newcommand{\var}[1]{\textnormal{Var}\bracket{#1}}
\newcommand{\set}[1]{\ensuremath{\left\{ #1 \right\}}}
\newcommand{\poly}{\mbox{\rm poly}}


%=====================================================
% Fill Out This Part With Your Own Information:
%=====================================================


\renewcommand{\thishomework}{4} %Homework number
\renewcommand{\thisname}{FIRST$_1$ LAST$_1$, FIRST$_2$ LAST$_2$, FIRST$_3$ LAST$_3$} % Enter your name here


\begin{document}

\thisheading

\subsection*{Homework Policy}
\begin{itemize}
\item You may consult all course materials (lecture notes, textbook, slides, etc.) while writing your solutions. No external resources are permitted.


\item \textbf{All submissions should be typeset in LaTeX, handwritten solutions are not accepted}. 

\item Remember to always \textbf{prove the correctness} of your algorithms and \textbf{analyze their running time} (or any other efficiency measure asked in the question).

\item Homework assignments are completed and submitted in fixed groups of 3 students. Only \textbf{one member of the group submits} the full solution on Canvas. The submission must include the names of all group members. The other two members do \textbf{not need to submit anything separately}.

\item You may discuss the problems with any of your classmates, even those outside your group. However, \textbf{each group must write and submit their solutions independently}.

\item Attempt the challenge problems only if you find them interesting. 
\end{itemize}



\subsection*{Acknowledgment (to be reviewed and edited as needed)}
\begin{tcolorbox}[colframe=black,colback=white,arc=5mm]
I, \textbf{Iris Martinez (bam322)} worked individually on this assignment. I used several youtube video lectures to help me better understand the material. I used NotebookLM to help me search and reference lecture materials and my notes.\textbf{I hereby affirm that all external assistance utilized in the preparation of this document has been properly acknowledged}.
\end{tcolorbox}

\finishline
\newpage


\newpage


\begin{problem}
	Suppose we have $n$ persons with their heights given in an array $P[1:n]$ and $n$ skis with heights given in an array $S[1:n]$. Design and analyze a greedy algorithm that in $O(n\log{n})$ time 
	assign a ski to each person so that the average difference between the height of a person and their assigned ski is minimized. The algorithm should output a permutation $\sigma$ of $\set{1,\ldots,n}$ with the meaning 
	that person $i$ is assigned the ski $\sigma(i)$ such that 
	\[
		\frac1n \cdot \sum_{i=1}^{n} |P[i] - S[\sigma(i)]|
	\]
	is minimized. 
	
	\medskip

\paragraph{Example:} If $P=[3,2,5,1]$ and $B=[5,7,2,9]$, we can output $\sigma = [2,1,4,3]$ with value
\[
	\frac{1}{4} \cdot  \sum_{i=1}^{4} |P[i]-B[\sigma(i)]| = \frac14 \cdot (|P[1]-B[2]| + |P[2] - B[1]| + |P[3] - B[4]| + |P[4]-B[3]|) = \frac{1}{4} \cdot ({4+3+4+1}) = 3, 
\]
which you can also check is the minimum value. 


\end{problem} 





\begin{solution}

%=================================================
    % 1. THE ALGORITHM %=====================================================
    \

    \textbf{Algorithm:}
    \begin{itemize}
        \item sort each array. This takes nlogn time. Now take the shortest person and match them to the shortest ski. Do this over until all skis are assigned.
    \end{itemize}




%=================================================
    % 2. PROOF OF CORRECTNESS %=====================================================
    \

    \textbf{Proof: } This must give the lowest average gap between person height and ski height because, given an order of skis where the shortest person was not matched with the shortest ski, then the gap between the shortest person and their assigned ski would be greater, therefore increasing the average gap by at least the amount improved by using that ski to match with another person.
    Formally:
    Assume an optimal pairing of people and skis A with an average difference of b, where you could turn A into the greedy assignment G with one swap. In pairing A, the average difference is $\frac{P_1 - S_2 + P_2 - S_1 + ... + P_n - S_n}{n}$. If we swapped this to the greedy solution, for the greedy solution to be as good or better we must prove this inequality: $|P_1 - S_2| + |P_2 - S_1| \ge |P_1 - S_1| + |P_2 - S_2|$ where the heights of $P_1 < P_2$ and the height of $S_1 < S_2$. The improvement gained by assigning $S_1$ to $P_2$ will make the match between $S_2$ and $P_1$ worst by the same amount or less than the amount of the improvement gained by the swap. Therefore using the greedy ordering is always at least as good or the same as a some other ordering, and the greedy approach is proven.




%=================================================
    % 3. RUN TIME ANALYSIS %=====================================================
    
    \textbf{Runtime: } Sorting an array takes nlogn time with mergesort. Running that twice gives 2nlogn. The pairing takes n time. Therefore in the limit the algorithim is $O(nlogn)$


\end{solution}




\newpage




\begin{problem}
    Given a set of jobs, each with a deadline and a processing time, schedule the jobs to minimize the maximum lateness, where lateness is defined as the time difference between a job’s completion and its deadline (if it’s completed after the deadline). Write an algorithm and prove its correctness.
\end{problem}






\begin{solution}

    %====================%
    %= 1. THE ALGORITHM =%
    %====================%
    \

    \textbf{Algorithm: }
    \

    To minimize the max lateness of jobs, we sort the jobs by deadline and always pick the job in order of earliest deadline. 



%=================================================
    % 2. PROOF OF CORRECTNESS %=====================================================
    \

    \textbf{Proof: }
        Let's assume there was a solution other than the greedy solution. This would mean there exists an occasion where we are doing a job with a later deadline before one with an earlier deadline. In this case, inverting that execution order to match the greedy solution will either not increase the max lateness(in the case that the max lateness occurs due to another job). We also know that it might give a better solution by decreasing the lateness of the event with the earlier deadline. The event with the later deadline, even if it becomes more late, necessarily cannot increase our maximum because the new lateness must necessarily be less than the lateness that the job with the earlier deadline was facing in the "optimal" solution. Therefore the greedy solution is optimal.





%=================================================
    % 3. RUN TIME ANALYSIS %=====================================================
    \


    \textbf{Runtime: }
    This only needs to run one sort then retrieve the items in order of the sort, so it runs in $O(nlogn)$ time


\end{solution}


\newpage



\begin{problem}
    Alice wants to throw a party and she is trying to decide who to invite. She has $n$ people to choose from, and she knows which pairs of these people know each other. She wants to invite as many people as possible, subject to two constraints:

\begin{itemize}
    \item For each guest, there should be at least five other guests that they already know.
\item For each guest, there should be at least five other guests that they don't already know.

\end{itemize}
Describe and analyze an algorithm that computes the largest possible number of guests Alice can invite, given a list of $n$ people and the list of pairs who know each other.
\end{problem}





\begin{solution}

%=================================================
    % 1. THE ALGORITHM %=====================================================
    \

    \textbf{Algorithm: }

    The setup of a list of people and a list of pairs that know each other can represent a graph where every node is a person and every edge is a pair of people that know each other. 

    Iterate through all nodes, calculating the number of outbound edges for every node and the number of "unknown" nodes as well (which is the same as the total number of nodes N - the number of outgoing edges).

    For any node we find where either the number of outbound edges or unknown edges is less than five, we add them to a queue for deletion,D.
    
    For every node in the deletion queue, find and remove it from the graph. Then recalculate outgoing edges for all nodes connected to the node. Recalculate unkown nodes for all not related nodes. If any nodes break the rules, add them to the deltion queue. Repeat until the deltion queue is empty. The number of nodes in the graph is the maximum number of invitable people.

%=================================================
    % 2. PROOF OF CORRECTNESS %=====================================================
    \

    \textbf{Proof: }
    This algorithm must be correct because it progressively trims people that don't meet the conditions until it arrives at a maximal set of them to invite. If there were a party larger than one given by this algorithm, this means the algorithm must remove someone wrongly. Because our algorithm runs the outgoing edges and unknown nodes calculation on each node, it cannot ever remove a node that isn't below the thresholds. Therefore, the algorithm cannot arrive at a set that is too small. We can prove this by assuming that the optimal party, O, has one more node than ours, P. Lets call the "mistakenly" removed node of P, x. Our algorithm evaluates each node to see if it should be removed by checking against the conditions. If a node x was removed, it must have lacked enough outgoing edges or unknown nodes. These values cannot increase with any future actions, and therefore deleting a node is always the correct move. Because this example node x is the first and only removal, a previous move could not have invalidated it. Because this is the case, every subsequent graph generated this way also is valid, and therefore every further removal of $x_n$ satisfies the same property of being inherently a valid or "correct" deletion.



%=================================================
    % 3. RUN TIME ANALYSIS %=====================================================
    \

    \textbf{Runtime: }
    The algorithm will run in $O(N^2)$ time. It will take $O(N^2)$ time to calculate the outgoing edges and unknown nodes values. In the worst case, we would cascade into a case where every single node would need to be removed, resulting in $N$ calculations for all $N$ nodes, giving a runtime of $O(2N^2)$ which simplifies to $O(N^2)$


\end{solution}

\newpage




\begin{problem}
A \textbf{walk} in a directed graph $G=(V,E)$ from a vertex $s$ to a vertex $t$, is a sequence of vertices $v_1,v_2,\ldots,v_k$ where $v_1 = s$ and $v_k=t$ such that for any $i < k$,  $(v_i,v_{i+1})$ is an edge in $G$. The \emph{length} of a walk
is defined as the number of vertices inside it minus one, i.e., the number of edges (so the walk $v_1,v_2,\ldots,v_k$ has length $k-1$). 

Note that the only difference 
of a walk with a \textbf{path} we defined in the course is that a walk can contain the same vertex (or edge) more than once, while a path consists of only distinct vertices and edges. 

Design and analyze an $O(n+m)$ time algorithm that given a directed graph $G=(V,E)$ and two vertices $s$ and $t$, outputs \emph{Yes} if there is a walk from $s$ to $t$ in $G$ \underline{whose length is even}, and \emph{No} otherwise. 


\end{problem}


\begin{solution}

%=================================================
    % 1. THE ALGORITHM %=====================================================
    \

    \textbf{Algorithm: }
    \

    We use a modified BFS with the following stucture. Rather than keeping track of visited nodes, we track visted nodes with even and odd parity. This is an array of length 2N where for all N vertices, there exists a vertex $N_odd$ and $N_even$. This allows us to model walks that reach a node in even or odd steps from the start node.
    When bfs exits (having an empty queue), then all the odd indices of the array contain true or false for which vertices can be reached from the origin vertex in an even number of steps. 
    If we return index $2E-1$ where E is the number of the target node, then we correctly give the true/false of an even walk from V to E
%=================================================
    % 2. PROOF OF CORRECTNESS %=====================================================
    \

    \textbf{Proof: }
    \

    This is true because we know BFS finds all vertices in a graph(if they are connected). By expanding the array to double the size, we are effectively making a new graph where there are twice as many nodes, one for a node reached with an even number of steps and another for the same node reached with an odd number of steps. When we traverse this graph starting at vertex V, we are only able to reach odd vertices, and all those that can be reached are pushed on to the stack. Then on the next step we can only reach even, and those are pushed onto the queue. We repeat this until we can no longer find any new vertices (find a vertex with an even or an odd walk). The output array will then contain true for all the connected nodes of this doubled graph. The connected nodes represent the idea of node x reached evenly and node x reached oddly,etc. Because bfs fully reaches all possible nodes in a graph, this discovers if there exists an even  walk between any vertex and the start vertex. 

%=================================================
    % 3. RUN TIME ANALYSIS %=====================================================
    \

    \textbf{Runtime: }

    This runs in O(N+M) time because it is basically just BFS of a graph that is larger than the original. This size increase scales it by a constant factor, so it simplifies to BFS, which runs in O(N+M) 
\end{solution}

\newpage





\begin{problem}
	We have an undirected graph $G=(V,E)$ and two vertices $s$ and $t$. Additionally, we are given a list of \emph{new} edges $F$ which do \emph{not} belong to the graph $G$. For any edge $f \in F$, 
	we define the graph $G_f$ as the graph obtained by adding the edge $f$ to the original graph $G$. 
	
	Design an algorithm that in time $O(n+m+|{F}|)$ finds the edge $f \in F$ such that 
	the distance from $s$ to $t$ in $G_f$ is minimized among all choices of an edge from $F$.  
\end{problem}



\begin{solution}

%=================================================
    % 1. THE ALGORITHM %=====================================================
    \

    \textbf{Algorithm: }

    Run BFS. Construct a bfs tree from vertex s. 
    Run BFS again. Construct a bfs tree from vertex t.

    Loop through every edge f in F. For any edge f, the distance between s and t in a graph that includes that edge is that same as the distance between the first vertex of f and s, and the distance between the second vertex in f and t, plus one.

    Because we have bfs trees from both ends, we get both of these distances for "free" in the form of the depths of all vertices.

    The expression $depth_s[u] + depth_t[v] + 1$ represents going from s to the tail end of the new edge and dist v from the tip of the new edge to the vertex t. The addition of one is to account for the new edge we are traversing in between. 

    our output is $newDepth = min(depth_s[u] + depth_t[v] + 1),depth_s(v)+depth_t[u], newDepth)$ for all f in F, newDepth is initialized to infty.

%=================================================
    % 2. PROOF OF CORRECTNESS %=====================================================
    
    \

    \textbf{Proof: }
    This must be true because for any unweighted graph, bfs always gives the shortest path between the root vertex and its children (where the depth of the BFS tree gives that distance). 
    If we have two such trees, we have the minimum distances from the origin and target nodes to every other node. 

    We can find what the change in distance would be by taking the distance between the input of an edge f (which we might call the vertex u) and the output of that edge(which we may call a vertex v) and adding one.

    This is true because we can view a new edge as shortcut of sorts. By figuring out how far it is to the entrance of the shortcut and how far it is to the destination from the exit of the shortcut, we know how long our total trip was. We try the edge forwards and backwards because we have an undirected graph. Therefore the minimum of our expression over all f edges gives the best edge to pick from F.  




%=================================================
    % 3. RUN TIME ANALYSIS %=====================================================
This runs in $O((2(m+n) + |F|)$ time because we run two BFS operations and then iterate over all F vertices doing constant time lookups. This simplifies down to $O(m + n + |F|)$ 


\end{solution}

\newpage




\begin{problem}
    Consider a directed graph $G$, where each edge is colored either red, white, or blue. A walk in $G$ is called a French flag walk if its sequence of edge colors is red, white, blue, red, white, blue, and so on. More formally, a walk $v_0 \rightarrow v_1 \rightarrow \cdots \rightarrow v_k$ is a French flag walk if, for every integer $i$, the edge $v_i \rightarrow v_{i+1}$ is red if $i \bmod 3=0$, white if $i \bmod 3=1$, and blue if $i \bmod 3=2$.

Describe an algorithm to find all vertices in $G$ that can be reached from a given vertex $v$ through a French flag walk. 
\end{problem}



\begin{solution}

%=================================================
    % 1. THE ALGORITHM %=====================================================
    \

    \textbf{Algorithm: }

    We will again modify BFS.

    This time, we will create a new graph with 3n nodes. For every node in the original graph, there are three nodes where each of the three represents reaching the original node via an edge of a given color.

    Now we will traverse this graph with bfs starting from the input node. However, we will only enqueue a node if it is of the correct color. So if we are at the node $A_white$, we will only enqueue any vertices of color blue. In order for this to run correctly, we must set the starting node to be $v_blue$, so that it looks for successors accessible via red edges. This represents the same as only traversing blue edges from A in a french flag walk. 

    When the algorithm is finished, all nodes in the tree are accessible via french flag walk starting at the input vertex.




%=================================================
    % 2. PROOF OF CORRECTNESS %=====================================================
    \textbf{Proof: }
    
    As in problem 4, all we are doing is making a new graph with more vertices that encodes not only the different nodes, but also the different kinds of edges that can reach that node. Because this is still just another graph, bfs still finds all reachable nodes starting from a vertex v. By modifying our traversal so that it ignores colors that are out of sequence, we make sure that all paths that the algorithm finds and puts into the tree are valid given our conditions. Therefore, because BFS is correct, this algorithm is also correct. 



%=================================================
    % 3. RUN TIME ANALYSIS %=====================================================
    \textbf{Proof: }
    This runs only one bfs and grows the graph by a constant factor. Therefore, this runs in $O(N+M)$ time where N is the number of vertices and M is the number of edges.


\end{solution}

\newpage




\begin{problem}
Suppose we are given both an undirected graph $G$ with weighted edges and a minimum spanning tree $T$ of $G$.
\begin{enumerate}
    \item [(a)]Describe an algorithm to update the minimum spanning tree when the weight of a single edge $e$ is decreased.
\item[(b)] Describe an algorithm to update the minimum spanning tree when the weight of a single edge $e$ is increased.

\end{enumerate}

\end{problem}



\begin{solution}

%=================================================
    % 1. THE ALGORITHM %=====================================================
    \

    \textbf{Algorithm A: } 
    If the edge e was in the MST, do nothing.

    If edge e was not in the MST do the following:

    Add e to the MST, T. Because T is the minimum spanning tree, any additional edge must form a cycle. 

    Run BFS on T starting at the vertex u (the input vertex of edge e). We do this because we know that this new edge is going to form a cycle starting at the vertex it originates from.
    
    We stop our BFS when we reach vertex v(the output vertex of edge e). At this point, we have found the full cycle we have created. We remove the heaviest edge in this cycle.

    We return the remaining tree, which is the new MST.


    
    \textbf{Algorithm B: } 

    If edge e was not in the MST, do nothing.

    If the edge was in the MST, we remove it. This splits the tree. We then check for all edges that connect the two halves of the tree. We keep the cheapest one.

    The way we check for edges that connect the two halves of the tree is by taking a list of of edges in each half, then iterating over the list of edges for all edges that connect vertices from one half to the other. 
%=================================================
    % 2. PROOF OF CORRECTNESS %=====================================================
    \textbf{Proof A: } 
    This algorithm works because adding a new edge to an mst necessarily creates a cycle. Because a cycle lets you visit all nodes along it, we can then remove the most expensive edge on that cycle to arrive at a new minimum spanning tree. The tree finds its cheapest configuration, and we don't lose the spanning property by removing just one edge from a cycle.

    \textbf{Proof B: }
    This works because if the cost of an edge not in the MST goes up, it makes no difference. However, if the cost of an edge in the mst goes up, it may mean that there is a better edge to include than it. 

    So we cut the mst in half. Each half is a mst of each subset of vertices. Now we find the cheapest connection between the two islands. This cheapest connection point will form from the two halves the MST of the whole. 


%=================================================
    % 3. RUN TIME ANALYSIS %=====================================================

    \textbf{Runtime A: }
    This algorithm runs in O(T) time, where T is the number of vertices in T. This is because at worst it only needs to iterate over all vertices in the tree once to figure out which to remove.

    \textbf{Runtime B: }
    This algorithm runs in O(E) time, where E is the number of edges in the whole graph. This is because in the worst case, we end up having to check every edge in the graph to find which form bridges between the two islands formed after we split the MST and keep track of the cheapest bridge to form the new MST.



\end{solution}

\newpage




\finishline

\begin{challenge}
Suppose you are given an arbitrary directed graph $G$ in which each edge is colored either red or blue, along with two special vertices $s$ and $t$.
\begin{enumerate}
    \item[(a)] Describe an algorithm that either computes a walk from $s$ to $t$ such that the pattern of red and blue edges along the walk is a palindrome, or correctly reports that no such walk exists.
\item[(b)] Describe an algorithm that either computes the shortest walk from $s$ to $t$ such that the pattern of red and blue edges along the walk is a palindrome, or correctly reports that no such walk exists.
	\end{enumerate}
\end{challenge}



\begin{solution}

%=================================================
    % 1. THE ALGORITHM %=====================================================





%=================================================
    % 2. PROOF OF CORRECTNESS %=====================================================





%=================================================
    % 3. RUN TIME ANALYSIS %=====================================================


\end{solution}

\newpage





\end{document}
